
---
title: "Reproducible Research - Programming Assigment 1"
author: "Javier Saravia"
date: "9 de septiembre de 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/Javier/Documents/2-D Especializacion en Ciencia de Datos/5 Reproducible Research/Unidad 2/Programming Asigment 1")
file1 <- read.csv("read.csv") # original file
file2 <- file1[complete.cases(file1),] # file without NA values
```

## TOTAL NUMBER OF STEPS TAKEN BY DAY

### 1 Calculate the total number of steps taken per day

```{r, echo=FALSE}
d1 <- duplicated(file2$date)
d2 <- file2$date[!d1]
days <- as.vector(d2) # 1 to 53

steps_sum <- as.vector(1:53)

for(i in 1:53)
{
  steps_sum[i] <- sum(file2$steps[file2$date == days[i]])  
}

steps_sum

```

### 2 Hist of total number of steps by day

```{r, echo=FALSE}
hist(steps_sum,
     main = 'Total steps by day',
     xlab = 'Steps',
     col="gray"
     )

```

### 3 Mean and median of the steps taken by day

```{r, echo=FALSE}

paste("The mean is",round(mean(steps_sum),0),"and the median is",round(median(steps_sum),0))

```

## WHAT IS THEAVERAGE DAILY ACTIVITY PATTERN?

### 1 Plot of the 5-minute interval and the average number of steps taken, average across all days

```{r, echo=FALSE}

steps_mean <- as.vector(1:53)

for(ii in 1:53)
{
  steps_mean[ii] <- round(mean(file2$steps[file2$date == days[ii]]),2)
}

steps_mean

hist(steps_mean,
     main = 'Mean steps by day',
     xlab = 'Steps',
     col="red"
)

```

### 2 Which 5-minutes interval, on average across all days in the dataset, contains the maximun númber of steps?

```{r, echo=FALSE}

interval1 <- duplicated(file2$interval)
interval <- file2$interval[!interval1] # 1 to 288

interval_mean <- 1:288

for(iii in 1:288)
{
  interval_mean[iii] = round(mean(file2$steps[file2$interval == interval[iii]]),2)
}   

interval_steps <- as.data.frame(cbind(interval,interval_mean))

interval_steps2 <- interval_steps[order(-interval_steps$interval_mean),]

head(interval_steps2)

```

## IMPUTING MISSING VALUES

### 1 Calculate and report the total number of missing values in the dataset

The NA values are in the STEPS Column

```{r, echo=FALSE}

print("Steps")
table(is.na(file1$steps))

print("Date")
table(is.na(file1$date))

print("Interval")
table(is.na(file1$interval))

```
NA values can be removed by using COMPLETE.CASES function

The code used for remove NA values is: file2 <- file1[complete.cases(file1),]


```{r, echo=FALSE}

file1 <- read.csv("read.csv") # original file
file2 <- file1[complete.cases(file1),] # file without NA values

paste("Rows before removing NA values: ",nrow(file1))

paste("Rows after removing NA values: ",nrow(file2))

```


### 2 Device a strategy for filling the NA values 

The missing (NA) values will be filled by the interval mean using this code:

```

steps_na <- 1:17568

for(s in 1:17568)

if (is.na(file1$steps[s])) 
  
  steps_na[s] = interval_steps2$interval_mean[interval_steps2$interval == file1$interval[s]]

else
  
  steps_na[s] = file1$steps[s]
  
```  

The resulting vector (steps_na) does not have NA vales:

```{r, echo=FALSE}

steps_na <- 1:17568

for(s in 1:17568) 
{
  if (is.na(file1$steps[s])) 
    { 
      steps_na[s] = interval_steps2$interval_mean[interval_steps2$interval == file1$interval[s]]
    } 
    else
    {
      steps_na[s] = file1$steps[s]
    }  
}

table(is.na(steps_na))

```

### 3 Create a new dataset that is equal to the original data set but with the missing values filled in

The new dataset has the same number of rows of the original dataset. Demostration:

```{r, echo=FALSE}

file3 <- cbind(file1,steps_na)

paste("Original dataset rows: ",nrow(file1))

paste("New dataset rows: ", nrow(file3))

```

### 4 Using the new dataset, make a histogram of the total number of steps taken by day and calculate and report the Mean and Median

The sum of steps by each days is:

```{r, echo=FALSE}

steps_sum <- as.vector(1:53)

for(vi in 1:53)
{
  steps_sum[vi] <- sum(file3$steps_na[file3$date == days[vi]])  
}

steps_sum

```

The histogram is:

```{r, echo=FALSE}

hist(steps_sum,
     main = 'Total steps by day',
     xlab = 'Steps',
     col="blue"
)

```

## ARE THERE DIFFERENCES IN ACTIVITY PATTERNS BETWEEN WEEKDAYS AND WEEKENDS?

### 1 Create a new variable in the dataset that indicate when a date is weekday or weekend

The function WEEKDAYS() can be used for create a new column that contain a 1 when the date is a weekday and a 2 then the date is a weekend

The new vector cointain the following values:

```{r, echo=FALSE}

week_day <- 1:17568

for(iiii in 1:17568)
{
  if(weekdays(as.Date(file3$date[iiii])) == "lunes") { week_day[iiii] = 1 }
  if(weekdays(as.Date(file3$date[iiii])) == "martes") { week_day[iiii] = 1 }
  if(weekdays(as.Date(file3$date[iiii])) == "miércoles") { week_day[iiii] = 1 }
  if(weekdays(as.Date(file3$date[iiii])) == "jueves") { week_day[iiii] = 1 }
  if(weekdays(as.Date(file3$date[iiii])) == "viernes") { week_day[iiii] = 1 }
  if(weekdays(as.Date(file3$date[iiii])) == "sábado") { week_day[iiii] = 2 }
  if(weekdays(as.Date(file3$date[iiii])) == "domingo") { week_day[iiii] = 2 }
}

table(week_day)

```

The new dataset is:

```{r, echo=FALSE}

file4 <- cbind(file3,week_day)

head(file4)

```

### 2 Plot of the average of steps by time interval and the average of steps by weekdays and weekends

The average of the steps by interval (using the last dataset created) only for weekdays is:

```{r, echo=FALSE}

interval_week_day <- 1:288
interval_week_end <- 1:288

for (i_day in 1:288)
{
  interval_week_day[i_day] = round(mean(file4$steps_na[file4$interval == interval[i_day] & file4$week_day[i_day] == 1]),2)
  
  interval_week_end[i_day] = round(mean(file4$steps_na[file4$interval == interval[i_day] & file4$week_day[i_day] == 2]),2)
}

hist(interval_week_day,
     main = 'Average steps by interval - weekday',
     xlab = 'Steps',
     col="purple"
)


```

* ¡Is not possible to generate a hist of steps average by interval for weekends cause the generation of NA values!

### 3 Average of steps by interval for weekday and weekend

The average of steps for weekdays and weekends are:

```{r, echo=FALSE}

paste("weekday steps average: ",round(mean(file4$steps_na[file4$week_day == 1]),2))

paste("weekend steps average: ",round(mean(file4$steps_na[file4$week_day == 2]),2))

```




 

